quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-07-01 paper-quollr.tex
Add the code
Updated the notation parameters

This function requires several parameters: the high-dimensional data (\texttt{highd\_data}), the emdedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{bin1}), the buffer amount as a proportion of data (\texttt{q}), and benchmark value to extract high density hexagons (\texttt{benchmark\_highdens}). The function returns an object that includes the scaled NLDR object (\texttt{nldr\_obj}), the hexagonal object (\texttt{hb\_obj}), the fitted model in both \(2\text{-}D\) (\texttt{model\_2d}), and \(p\text{-}D\) (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

\begin{verbatim}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  bin1 = 15, 
  q = 0.1, 
  benchmark_highdens = 5)
\end{verbatim}

@@

The configurations of a hexagonal grid are determined by the number of bins and the bin width in each direction. The function \texttt{calc\_bins\_y()} is used for this purpose. This function accepts an object containing scaled NLDR data in the first and second columns, along with numeric vectors that represent the limits of the original NLDR data, the number of bins along the x-axis (\texttt{bin1}), and the buffer amount as a proportion (\texttt{q}).

\begin{verbatim}
bin_configs <- calc_bins_y(
  nldr_obj = scurve_umap_obj, 
  bin1 = 15, 
  q = 0.1)

@@

\begin{verbatim}
#> $bin2
#> [1] 16
#> 
#> $a1
#> [1] 0.08326136
#> 
#> $a2
#> [1] 0.07210645
\end{verbatim}

@@

Points are allocated to bins based on the nearest centroid. The hexagonal binning algorithm can be executed using the \texttt{hex\_binning()} function, or its components can be run separately for added flexibility. The parameters used within \texttt{hex\_binning()} include an object containing scaled NLDR data in the first and second columns, along with numeric vectors that represent the limits of the original NLDR data (\texttt{nldr\_obj}), the number of bins along the x-axis (\texttt{bin1}), and the buffer amount as a proportion of the data (\texttt{q}). The output is an object of the \texttt{hex\_bin\_obj} class, which contains the bin widths in each direction (\texttt{a1}, \texttt{a2}), the number of bins in each direction (\texttt{bins}), the coordinates of the hexagonal grid starting point (\texttt{start\_point}), the details of bin centroids (\texttt{centroids}), the coordinates of bins (\texttt{hex\_poly}), NLDR components with their corresponding hexagon IDs (\texttt{data\_hb\_id}), hex bins with their corresponding standardized counts (\texttt{std\_cts}), the total number of bins (\texttt{tot\_bins}), the number of non-empty bins (\texttt{non\_bins}), and the points within each hexagon (\texttt{pts\_bins}).

\begin{verbatim}
hb_obj <- hex_binning(
  nldr_obj = scurve_umap_obj, 
  bin1 = 15, 
  q = 0.1)
\end{verbatim}

@@

The coordinate limits of the embedding (\texttt{lim1} and \texttt{lim2}) are used to compute the aspect ratio between the two axes, which informs vertical spacing. The function then calls \texttt{calc\_bins\_y()}, a helper function that determines the appropriate number of hexagonal rows (\texttt{bin2}) and the width of each hexagon (\texttt{a1}) given the specified number of bins along the x-axis (\texttt{bin1}) and buffer (\texttt{q}).

@@

Finally, a tibble is returned containing a unique hexagon ID (\texttt{hexID}) along with the corresponding x and y centroid coordinates (\texttt{c\_x}, \texttt{c\_y}), which define the layout of the hexagonal grid over the \(2-\text{D}\) space.

\begin{verbatim}
all_centroids_df <- gen_centroids(
  nldr_obj = scurve_umap_obj, 
  bin1 = 15, 
  q = 0.1
  )

@@

\begin{verbatim}
#> # A tibble: 240 x 3
#>    hexID     c_x     c_y
#>    <int>   <dbl>   <dbl>
#>  1     1 -0.1    -0.0892
#>  2     2 -0.0167 -0.0892
#>  3     3  0.0665 -0.0892
#>  4     4  0.150  -0.0892
#>  5     5  0.233  -0.0892
#>  6     6  0.316  -0.0892
#>  7     7  0.400  -0.0892
#>  8     8  0.483  -0.0892
#>  9     9  0.566  -0.0892
#> 10    10  0.649  -0.0892
#> # i 230 more rows
\end{verbatim}

@@

For each centroid, six vertices are computed and assigned a polygon ID (\texttt{hex\_poly\_id}) corresponding to the centroid's \texttt{hexID}. These points are aggregated into a tibble that includes the polygon ID (\texttt{hex\_poly\_id}) and the respective x (\texttt{x}) and y (\texttt{y}) coordinates for all hexagon corners.

@@

Both the embedding coordinates and the centroid coordinates are converted to matrices to facilitate distance computations. The function uses the \texttt{proxy::dist()} method to compute a pairwise Euclidean distance matrix between all NLDR points and all centroids. For each NLDR point, the function identifies the index of the centroid with the smallest distance representing the closest hexagon---and assigns the corresponding hexagon ID (\texttt{hexID}) to the point.

The result is a data frame of the scaled \(2-\text{D}\) embedding with an additional \texttt{hexID} column, indicating the hexagonal bin to which each point belongs.

@@

\begin{verbatim}
#> # A tibble: 5 x 4
#>    emb1  emb2    ID hexID
#>   <dbl> <dbl> <int> <int>
#> 1 0.707 0.839     1   205
#> 2 0.231 0.401     2   109
#> 3 0.232 0.215     3    65
#> 4 0.790 0.564     4   146
#> 5 0.761 0.551     5   146
\end{verbatim}

@@

The function begins by grouping the data by \texttt{hexID} and counting the number of NLDR points falling within each bin. These raw counts are stored as \texttt{bin\_counts}. To enable comparisons across bins with varying densities, the function then standardizes these counts by dividing each bin's count by the maximum count across all bins. This yields a normalized metric, \texttt{std\_counts}, ranging from 0 to 1.

\begin{verbatim}
std_df <- compute_std_counts(
  scaled_nldr_hexid = umap_hex_id
  )

@@

\begin{verbatim}
#> # A tibble: 5 x 3
#>   hexID bin_counts std_counts
#>   <int>      <int>      <dbl>
#> 1    21         12     0.16  
#> 2    22         17     0.227 
#> 3    23         22     0.293 
#> 4    24         10     0.133 
#> 5    25          7     0.0933
\end{verbatim}

@@

The function first groups the input data by \texttt{hexID}, which represents the hexagon label associated with each point in the \(2-\text{D}\) layout. Within each group, it collects the \texttt{ID}s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, \texttt{pts\_list}, contains all point identifiers associated with that hexagon.

@@

\begin{verbatim}
#> # A tibble: 5 x 2
#>   hexID pts_list  
#>   <int> <list>    
#> 1    21 <int [12]>
#> 2    22 <int [17]>
#> 3    23 <int [22]>
#> 4    24 <int [10]>
#> 5    25 <int [7]>
\end{verbatim}

@@

This function begins by arranging the \texttt{counts\_data} by \texttt{hexID} to ensure consistent ordering. It then performs a full join with \texttt{centroids\_data}, aligning hexagon IDs between the two datasets to incorporate both hexagonal bin centroids and count metrics. After merging, the function handles missing values in the count columns: any \texttt{NA} values in \texttt{std\_counts} or \texttt{bin\_counts} are replaced with zeros. This ensures that hexagons with no assigned data points are retained in the output, with zero values for count-related fields. The resulting data frame contains the full set of hexagon centroids along with associated bin counts and standardized counts.

@@

\begin{verbatim}
#>   hexID         c_x         c_y bin_counts std_counts
#> 1     1 -0.10000000 -0.08923607          0          0
#> 2     2 -0.01673864 -0.08923607          0          0
#> 3     3  0.06652271 -0.08923607          0          0
#> 4     4  0.14978407 -0.08923607          0          0
#> 5     5  0.23304543 -0.08923607          0          0
\end{verbatim}

@@

The \texttt{find\_low\_dens\_hex()} function identifies hexagons with low point densities, considering the densities of their neighboring bins as well. The \texttt{find\_low\_dens\_hex()} function relies on a helper, \texttt{compute\_mean\_density\_hex()}, which iterates over all hexagons and computes the average density across neighbors based on their \texttt{hexID} and a defined number of bins along the x-axis (\texttt{bin1}). The hexagonal layout assumes a fixed grid structure, so neighbor IDs are computed by positional offsets.

\begin{verbatim}
find_low_dens_hex(
  centroids_data = df_bin_centroids, 
  bin1 = 15, 
  benchmark_mean_dens = 0.05
)
\end{verbatim}

\begin{verbatim}
#>  [1]   1   2   3   4   5  11  12  13  14  15  16  17  18  19  30  31  32  33  46
#> [20]  47  61  75  76  90 105 120 135 150 151 166 181 182 196 197 198 211 212 213
#> [39] 214 215 225 226 227 228 229 230 231 232 233 234 235 236 237 238 239 240
\end{verbatim}

@@

\begin{verbatim}
df_bin_centroids <- df_bin_centroids |>
  dplyr::filter(bin_counts > 10)

@@

\begin{verbatim}
#> # A tibble: 5 x 8
#>   hexID      x1    x2    x3         x4        x5       x6        x7
#>   <int>   <dbl> <dbl> <dbl>      <dbl>     <dbl>    <dbl>     <dbl>
#> 1    21 -0.992   1.91  1.11 -0.000427   0.000624  0.00749  0.00105 
#> 2    22 -0.906   1.93  1.41 -0.0000183  0.00331  -0.0204  -0.000363
#> 3    23 -0.680   1.93  1.72 -0.000810  -0.00259  -0.00449  0.00153 
#> 4    24 -0.272   1.93  1.96  0.00251    0.00668  -0.0460   0.00128 
#> 5    25  0.0760  1.93  2.00  0.00876    0.00447   0.00851 -0.00195
\end{verbatim}

@@

>>>>

This function requires several parameters: the high-dimensional data (\texttt{highd\_data}), the emdedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{b1}), the buffer amount as a proportion of data (\texttt{q}), and benchmark value to extract high density hexagons (\texttt{benchmark\_highdens}). The function returns an object that includes the scaled NLDR object (\texttt{nldr\_obj}), the hexagonal object (\texttt{hb\_obj}), the fitted model in both \(2\text{-}D\) (\texttt{model\_2d}), and \(p\text{-}D\) (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

\begin{verbatim}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 15, 
  q = 0.1, 
  benchmark_highdens = 5)
\end{verbatim}

@@

The configurations of a hexagonal grid are determined by the number of bins and the bin width in each direction. The function \texttt{calc\_bins\_y()} is used for this purpose. This function accepts an object containing scaled NLDR data in the first and second columns, along with numeric vectors that represent the limits of the original NLDR data, the number of bins along the x-axis (\texttt{b1}), and the buffer amount as a proportion (\texttt{q}).

\begin{verbatim}
bin_configs <- calc_bins_y(
  nldr_obj = scurve_umap_obj, 
  b1 = 15, 
  q = 0.1)

@@

\begin{verbatim}
#> $b2
#> [1] 16
#> 
#> $a1
#> [1] 0.08326136
#> 
#> $a2
#> [1] 0.07210645
\end{verbatim}

@@

Points are allocated to bins based on the nearest centroid. The hexagonal binning algorithm can be executed using the \texttt{hex\_binning()} function, or its components can be run separately for added flexibility. The parameters used within \texttt{hex\_binning()} include an object containing scaled NLDR data in the first and second columns, along with numeric vectors that represent the limits of the original NLDR data (\texttt{nldr\_obj}), the number of bins along the x-axis (\texttt{b1}), and the buffer amount as a proportion of the data (\texttt{q}). The output is an object of the \texttt{hex\_bin\_obj} class, which contains the bin widths in each direction (\texttt{a1}, \texttt{a2}), the number of bins in each direction (\texttt{bins}), the coordinates of the hexagonal grid starting point (\texttt{start\_point}), the details of bin centroids (\texttt{centroids}), the coordinates of bins (\texttt{hex\_poly}), NLDR components with their corresponding hexagon IDs (\texttt{data\_hb\_id}), hex bins with their corresponding standardized counts (\texttt{std\_cts}), the total number of bins (\texttt{tot\_bins}), the number of non-empty bins (\texttt{non\_bins}), and the points within each hexagon (\texttt{pts\_bins}).

\begin{verbatim}
hb_obj <- hex_binning(
  nldr_obj = scurve_umap_obj, 
  b1 = 15, 
  q = 0.1)
\end{verbatim}

@@

The coordinate limits of the embedding (\texttt{lim1} and \texttt{lim2}) are used to compute the aspect ratio between the two axes, which informs vertical spacing. The function then calls \texttt{calc\_bins\_y()}, a helper function that determines the appropriate number of hexagonal rows (\texttt{bin2}) and the width of each hexagon (\texttt{a1}) given the specified number of bins along the x-axis (\texttt{b1}) and buffer (\texttt{q}).

@@

Finally, a tibble is returned containing a unique hexagon ID (\texttt{h}) along with the corresponding x and y centroid coordinates (\texttt{c\_x}, \texttt{c\_y}), which define the layout of the hexagonal grid over the \(2-\text{D}\) space.

\begin{verbatim}
all_centroids_df <- gen_centroids(
  nldr_obj = scurve_umap_obj, 
  b1 = 15, 
  q = 0.1
  )

@@

\begin{verbatim}
#> # A tibble: 240 x 3
#>        h     c_x     c_y
#>    <int>   <dbl>   <dbl>
#>  1     1 -0.1    -0.0892
#>  2     2 -0.0167 -0.0892
#>  3     3  0.0665 -0.0892
#>  4     4  0.150  -0.0892
#>  5     5  0.233  -0.0892
#>  6     6  0.316  -0.0892
#>  7     7  0.400  -0.0892
#>  8     8  0.483  -0.0892
#>  9     9  0.566  -0.0892
#> 10    10  0.649  -0.0892
#> # i 230 more rows
\end{verbatim}

@@

For each centroid, six vertices are computed and assigned a polygon ID (\texttt{hex\_poly\_id}) corresponding to the centroid's \texttt{h}. These points are aggregated into a tibble that includes the polygon ID (\texttt{hex\_poly\_id}) and the respective x (\texttt{x}) and y (\texttt{y}) coordinates for all hexagon corners.

@@

Both the embedding coordinates and the centroid coordinates are converted to matrices to facilitate distance computations. The function uses the \texttt{proxy::dist()} method to compute a pairwise Euclidean distance matrix between all NLDR points and all centroids. For each NLDR point, the function identifies the index of the centroid with the smallest distance representing the closest hexagon---and assigns the corresponding hexagon ID (\texttt{h}) to the point.

The result is a data frame of the scaled \(2-\text{D}\) embedding with an additional \texttt{h} column, indicating the hexagonal bin to which each point belongs.

@@

\begin{verbatim}
#> # A tibble: 5 x 4
#>    emb1  emb2    ID     h
#>   <dbl> <dbl> <int> <int>
#> 1 0.707 0.839     1   205
#> 2 0.231 0.401     2   109
#> 3 0.232 0.215     3    65
#> 4 0.790 0.564     4   146
#> 5 0.761 0.551     5   146
\end{verbatim}

@@

The function begins by grouping the data by \texttt{h} and counting the number of NLDR points falling within each bin. These raw counts are stored as \texttt{n\_h}. To enable comparisons across bins with varying densities, the function then standardizes these counts by dividing each bin's count by the maximum count across all bins. This yields a normalized metric, \texttt{w\_h}, ranging from 0 to 1.

\begin{verbatim}
std_df <- compute_std_counts(
  scaled_nldr_h = umap_hex_id
  )

@@

\begin{verbatim}
#> # A tibble: 5 x 3
#>       h   n_h    w_h
#>   <int> <int>  <dbl>
#> 1    21    12 0.0024
#> 2    22    17 0.0034
#> 3    23    22 0.0044
#> 4    24    10 0.002 
#> 5    25     7 0.0014
\end{verbatim}

@@

The function first groups the input data by \texttt{h}, which represents the hexagon label associated with each point in the \(2-\text{D}\) layout. Within each group, it collects the \texttt{ID}s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, \texttt{pts\_list}, contains all point identifiers associated with that hexagon.

@@

\begin{verbatim}
#> # A tibble: 5 x 2
#>       h pts_list  
#>   <int> <list>    
#> 1    21 <int [12]>
#> 2    22 <int [17]>
#> 3    23 <int [22]>
#> 4    24 <int [10]>
#> 5    25 <int [7]>
\end{verbatim}

@@

This function begins by arranging the \texttt{counts\_data} by \texttt{h} to ensure consistent ordering. It then performs a full join with \texttt{centroids\_data}, aligning hexagon IDs between the two datasets to incorporate both hexagonal bin centroids and count metrics. After merging, the function handles missing values in the count columns: any \texttt{NA} values in \texttt{w\_h} or \texttt{n\_h} are replaced with zeros. This ensures that hexagons with no assigned data points are retained in the output, with zero values for count-related fields. The resulting data frame contains the full set of hexagon centroids along with associated bin counts and standardized counts.

@@

\begin{verbatim}
#>   h         c_x         c_y n_h w_h
#> 1 1 -0.10000000 -0.08923607   0   0
#> 2 2 -0.01673864 -0.08923607   0   0
#> 3 3  0.06652271 -0.08923607   0   0
#> 4 4  0.14978407 -0.08923607   0   0
#> 5 5  0.23304543 -0.08923607   0   0
\end{verbatim}

@@

The \texttt{find\_low\_dens\_hex()} function identifies hexagons with low point densities, considering the densities of their neighboring bins as well. The \texttt{find\_low\_dens\_hex()} function relies on a helper, \texttt{compute\_mean\_density\_hex()}, which iterates over all hexagons and computes the average density across neighbors based on their \texttt{h} and a defined number of bins along the x-axis (\texttt{b1}). The hexagonal layout assumes a fixed grid structure, so neighbor IDs are computed by positional offsets.

\begin{verbatim}
find_low_dens_hex(
  centroids_data = df_bin_centroids, 
  b1 = 15, 
  benchmark_mean_dens = 0.05
)
\end{verbatim}

\begin{verbatim}
#>   [1]   1   2   3   4   5   6   7   8   9  10  11  12  13  14  15  16  17  18
#>  [19]  19  20  21  22  23  24  25  26  27  28  29  30  31  32  33  34  35  36
#>  [37]  37  38  39  40  41  42  43  44  45  46  47  48  49  50  51  52  53  54
#>  [55]  55  56  57  58  59  60  61  62  63  64  65  66  67  68  69  70  71  72
#>  [73]  73  74  75  76  77  78  79  80  81  82  83  84  85  86  87  88  89  90
#>  [91]  91  92  93  94  95  96  97  98  99 100 101 102 103 104 105 106 107 108
#> [109] 109 110 111 112 113 114 115 116 117 118 119 120 121 122 123 124 125 126
#> [127] 127 128 129 130 131 132 133 134 135 136 137 138 139 140 141 142 143 144
#> [145] 145 146 147 148 149 150 151 152 153 154 155 156 157 158 159 160 161 162
#> [163] 163 164 165 166 167 168 169 170 171 172 173 174 175 176 177 178 179 180
#> [181] 181 182 183 184 185 186 187 188 189 190 191 192 193 194 195 196 197 198
#> [199] 199 200 201 202 203 204 205 206 207 208 209 210 211 212 213 214 215 216
#> [217] 217 218 219 220 221 222 223 224 225 226 227 228 229 230 231 232 233 234
#> [235] 235 236 237 238 239 240
\end{verbatim}

@@

\begin{verbatim}
df_bin_centroids <- df_bin_centroids |>
  dplyr::filter(n_h > 10)

@@

\begin{verbatim}
#> # A tibble: 5 x 8
#>       h      x1    x2    x3         x4        x5       x6        x7
#>   <int>   <dbl> <dbl> <dbl>      <dbl>     <dbl>    <dbl>     <dbl>
#> 1    21 -0.992   1.91  1.11 -0.000427   0.000624  0.00749  0.00105 
#> 2    22 -0.906   1.93  1.41 -0.0000183  0.00331  -0.0204  -0.000363
#> 3    23 -0.680   1.93  1.72 -0.000810  -0.00259  -0.00449  0.00153 
#> 4    24 -0.272   1.93  1.96  0.00251    0.00668  -0.0460   0.00128 
#> 5    25  0.0760  1.93  2.00  0.00876    0.00447   0.00851 -0.00195
\end{verbatim}

@@

\begin{verbatim}
design <- gen_design(n_right = 6, ncol_right = 2)

plot_rmse_layouts(plots = list(error_plot_limb, nldr1, 
                               nldr2, nldr3, nldr4, 
                               nldr5, nldr6), design = design)
\end{verbatim}
